---
title: "Johns Hopkins Hospital Sequencing Frequency Report"
author: Elgin Akin
date: now
execute: 
  echo: false
format: 
  html:
    page-layout: full
    code-fold: true
    embed-resources: true
    toc: true
---

Most recent run: **IV25Run16** Not shown in plots below due to ambiguous date

```{r}
#| echo: false
#| output: false 

library(tidyverse)
library(RColorBrewer)
library(plotly)

#data <- read_tsv("../reports/report.tsv")
data <- read_tsv("reports/report.tsv")

data_filtered <- data %>% 
  dplyr::filter(
    subtype %in% c("H1N1", "H3N2", "Victoria"), 
    date >= as.Date("2023-01-01") & date <= as.Date("2025-12-31")
  ) %>%
  dplyr::mutate(date_run = paste(date, sequencing_run, sep = " | ")) 

# Ensure date_run is ordered chronologically
data_filtered <- data_filtered %>%
  mutate(date = as.Date(date)) %>%  # Ensure the date is in Date format
  arrange(date) %>%  # Sort data by date
  mutate(date_run = factor(date_run, levels = unique(date_run)))  # Create factor with ordered levels

```


## Subtype plot

```{r}
#| echo: FALSE
#| output: FALSE 
#| column: screen-inset-shaded 

# Define Dark2 palette for the subtypes
dark2_colors <- brewer.pal(n = 3, name = "Dark2")

# Plot stacked bar chart (showing each date-run pair separately)
subtype_frequency = ggplot(data_filtered, aes(x = factor(date_run, levels = unique(date_run)), fill = subtype)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = dark2_colors) +  # Apply Dark2 palette
  labs(title = "Johns Hopkins Hospital System - Subtype Frequency",
       x = "Sequence Date | Sequencing Run #",
       y = "Number of Sequences",
       fill = "Subtype") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))  +
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(vjust = 0.5))

```

```{r}
#| column: screen-inset-shaded 
ggplotly(subtype_frequency)
#ggsave(filename = paste0("../figures/subtype_frequency.png"), plot = subtype_frequency, width = 10, height = 4)
```

# Clade and Subclade Frequency (depreciated)
```{r}
#| echo: false
#| output: false 
library(tidyverse)

# Define the clade and subclade color palettes
clade_colors <- c(
  # H1N1
  "6B.1A.5a.2a" = "#a0240e", 
  "6B.1A.5a.2a.1" = "#6265cd", 
  # H3N2
  "3C.2a1b.2a.2a.1b" = "#A4BED5FF",
  "3C.2a1b.2a.2a.3a.1" = "#CC4C24FF", 
  "3C.2a1b.2a.2b" = "#32373AFF", 
  # IBV
  "V1A" = "#FF4500", 
  "V1A.3a.2" = "#00CED1"
)

subclade_colors <- c(
  # H1N1
  "C.1" = "#9A133DFF", 
  "C.1.1" = "#B93961FF", 
  "C.1.7" = "#D8527CFF", 
  "C.1.7.2" = "#F28AAAFF", 
  "C.1.8" = "#F9B4C9FF", 
  "C.1.9" = "#F9E0E8FF", 
  "D" = "#1A318BFF", 
  "D.1" = "#4060C8FF", 
  "D.2" = "#6996E3FF", 
  "D.3" = "#A1C2EDFF", 
  "D.4" = "#C5DAF6FF",
  # IBV
  "A" = "#FF6347", 
  "C.3" = "#00CED1", 
  "C.5" = "#9400D3", 
  "C.5.1" = "#FF1493", 
  "C.5.5" = "#FF8C00", 
  "C.5.6" = "#B0E0E6", 
  "C.5.7" = "#808000", 
  # H3N2
  "G.1.1.2" = "#00496FFF", 
  "G.2" = "#0F85A0FF", 
  "J" = "#EDD746FF", 
  "J.1" = "#ED8B00FF", 
  "J.2" = "#DD4124FF"
)

# Define flu season start date
flu_season_start <- as.Date("2024-10-01")

# Function to generate plots for a given subtype
generate_plots <- function(df, subtype_name) {
  df_filtered <- df %>%
    filter(
      subtype == subtype_name,
      date >= as.Date("2023-01-01") & date <= as.Date("2025-12-31"),
      !is.na(clade) & clade != "",
      !is.na(subclade) & subclade != ""
    ) %>%
    mutate(date_run = paste(date, sequencing_run, sep = " | ")) %>%  # Create a unique date-run label
    arrange(date) %>%  # Ensure data is sorted by date
    mutate(date_run = factor(date_run, levels = unique(date_run)))  # Set factor levels in time order

  # Clade Frequency Plot
  clade_plot <- ggplot(df_filtered, aes(x = date_run, fill = clade)) +
    geom_bar(position = "stack") +
    scale_fill_manual(values = clade_colors) +  # Apply clade color palette
    labs(title = paste(subtype_name, "Clade Frequency"),
         x = "Date | Sequencing Run",
         y = "Frequency",
         fill = "Clade") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
    geom_vline(xintercept = which(levels(df_filtered$date_run) == as.character(flu_season_start)), 
               linetype = "dotted", color = "red", size = 1) +
    geom_text(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5))

  # Subclade Frequency Plot
  subclade_plot <- ggplot(df_filtered, aes(x = date_run, fill = subclade)) +
    geom_bar(position = "stack") +
    scale_fill_manual(values = subclade_colors) +  # Apply subclade color palette
    labs(title = paste(subtype_name, "Subclade Frequency"),
         x = "Date | Sequencing Run",
         y = "Frequency",
         fill = "Subclade") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
    geom_vline(xintercept = which(levels(df_filtered$date_run) == as.character(flu_season_start)), 
               linetype = "dotted", color = "red", size = 1) +
    geom_text(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5))

  #ggsave(filename = paste0("../figures/", subtype_name, "_clade_plot.png"), plot = clade_plot, width = 10, height = 4)
  #ggsave(filename = paste0("../figures/", subtype_name, "_subclade_plot.png"), plot = subclade_plot, width = 10, height = 4)

  return(list(clade_plot = clade_plot, subclade_plot = subclade_plot))
}

# Generate plots for each subtype
h3n2_plots <- generate_plots(data, "H3N2")
h1n1_plots <- generate_plots(data, "H1N1")
victoria_plots <- generate_plots(data, "Victoria")
```

# Clade and Subclade Frequency
```{r}


```

# Clade and Subclade Frequency v3 

```{r}
library(RColorBrewer)

### Begin generating color palettes 

generate_palette <- function(n, palette) {
  max_colors <- brewer.pal.info[palette, "maxcolors"]
  base_colors <- brewer.pal(min(n, max_colors), palette)
  if (n <= max_colors) {
    return(base_colors)
  } else {
    return(colorRampPalette(base_colors)(n))
  }
}

# Assign palettes per subtype
clade_palette_map <- list(
  "H1N1"    = "Dark2",
  "H3N2"    = "Set2",
  "Victoria"= "Set1"
)

subclade_palette_map <- list(
  "H1N1"    = "Dark2",
  "H3N2"    = "PairSet2ed",
  "Victoria"= "Set1"
)

# Generate clade colors for a given subtype
get_clade_colors <- function(df, subtype) {
  palette <- clade_palette_map[[subtype]]
  
  clades <- df %>%
    filter(subtype == subtype, !is.na(`clade`), clade != "") %>%
    pull(clade) %>% unique() %>% sort()
  
  colors <- generate_palette(length(clades), palette)
  names(colors) <- clades
  colors
}

# Generate subclade colors for a given subtype
get_subclade_colors <- function(df, subtype) {
  palette <- subclade_palette_map[[subtype]]
  
  subclades <- df %>%
    filter(subtype == subtype, !is.na(subclade), subclade != "") %>%
    pull(subclade) %>% unique() %>% sort()
  
  colors <- generate_palette(length(subclades), palette)
  names(colors) <- subclades
  colors
}

### Begin generating plots 

generate_plots <- function(df, subtype_name) {
  
  df_filtered <- df %>%
    filter(
      subtype == subtype_name,
      date >= as.Date("2023-01-01") & date <= as.Date("2025-12-31"),
      !is.na(clade), clade != "",
      !is.na(subclade), subclade != ""
    ) %>%
    mutate(date_run = paste(date, sequencing_run, sep = " | ")) %>%
    arrange(date) %>%
    mutate(date_run = factor(date_run, levels = unique(date_run)))

  # *** NEW: subtype-specific palettes ***
  clade_colors    <- get_clade_colors(df_filtered, subtype_name)
  subclade_colors <- get_subclade_colors(df_filtered, subtype_name)

  # Clade Frequency Plot
  clade_plot <- ggplot(df_filtered, aes(x = date_run, fill = clade)) +
    geom_bar(position = "stack") +
    geom_text(stat = "count", aes(label = after_stat(count)), 
              position = position_stack(vjust = 0.5), 
              color = "white", fontface = "bold", size = 3) +
    scale_fill_manual(values = clade_colors) +
    labs(title = paste(subtype_name, "Clade Frequency"),
         x = "Date | Sequencing Run",
         y = "Frequency",
         fill = "Clade") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

  # Subclade Frequency Plot
  subclade_plot <- ggplot(df_filtered, aes(x = date_run, fill = subclade)) +
    geom_bar(position = "stack") +
    geom_text(stat = "count", aes(label = after_stat(count)), 
              position = position_stack(vjust = 0.5), 
              color = "white", fontface = "bold", size = 3) +
    scale_fill_manual(values = subclade_colors) +
    labs(title = paste(subtype_name, "Subclade Frequency"),
         x = "Date | Sequencing Run",
         y = "Frequency",
         fill = "Subclade") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))

  return(list(clade_plot = clade_plot, subclade_plot = subclade_plot))
}

# Generate plots for each subtype
h3n2_plots <- generate_plots(data, "H3N2")
h1n1_plots <- generate_plots(data, "H1N1")
victoria_plots <- generate_plots(data, "Victoria")
```

## H3N2
```{r}
#| column: screen-inset-shaded
ggplotly(h3n2_plots$clade_plot)
ggplotly(h3n2_plots$subclade_plot)
```

## H1N1
```{r}
#| column: screen-inset-shaded
ggplotly(h1n1_plots$clade_plot)
ggplotly(h1n1_plots$subclade_plot)
```

## B/Victoria
```{r}
#| column: screen-inset-shaded
ggplotly(victoria_plots$clade_plot)
ggplotly(victoria_plots$subclade_plot)
```

# Interactive Tables

## Total Clade and Subclade Summary by Subtype

```{r}
# Table 1: Overall clade/subclade summary by subtype
clade_subclade_summary <- data %>%
  filter(subtype %in% c("H1N1", "H3N2", "Victoria")) %>%
  group_by(subtype) %>%
  summarise(
    total_samples = n(),
    unique_clades = n_distinct(clade[!is.na(clade) & clade != ""]),
    unique_subclades = n_distinct(subclade[!is.na(subclade) & subclade != ""]),
    samples_with_clade = sum(!is.na(clade) & clade != ""),
    samples_with_subclade = sum(!is.na(subclade) & subclade != ""),
    clade_coverage = round((samples_with_clade / total_samples) * 100, 1),
    subclade_coverage = round((samples_with_subclade / total_samples) * 100, 1),
    .groups = 'drop'
  )

datatable(
  clade_subclade_summary,
  caption = "Clade and Subclade Summary by Subtype",
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE
  )) %>%
  formatStyle(columns = c('clade_coverage', 'subclade_coverage'), 
              backgroundColor = styleInterval(c(50, 80), c('lightcoral', 'lightyellow', 'lightgreen')))
```

## Seasonal Clade and Subclade Summary (October-September Seasons)

```{r}
# Table 2: Seasonal summary (Oct-Sep seasons)
seasonal_summary <- data %>%
  filter(
    subtype %in% c("H1N1", "H3N2", "Victoria"),
    !is.na(date)
  ) %>%
  mutate(
    # Create flu season (Oct-Sep)
    flu_season = case_when(
      month(date) >= 10 ~ paste0(year(date), "-", year(date) + 1),
      month(date) < 10 ~ paste0(year(date) - 1, "-", year(date)),
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(flu_season)) %>%
  group_by(flu_season, subtype) %>%
  summarise(
    total_samples = n(),
    unique_clades = n_distinct(clade[!is.na(clade) & clade != ""]),
    unique_subclades = n_distinct(subclade[!is.na(subclade) & subclade != ""]),
    samples_with_clade = sum(!is.na(clade) & clade != ""),
    samples_with_subclade = sum(!is.na(subclade) & subclade != ""),
    clade_coverage = round((samples_with_clade / total_samples) * 100, 1),
    subclade_coverage = round((samples_with_subclade / total_samples) * 100, 1),
    .groups = 'drop'
  ) %>%
  arrange(flu_season, subtype)

datatable(
  seasonal_summary,
  caption = "Seasonal Clade and Subclade Summary (October-September Seasons)",
  extensions = 'Buttons',
  filter = list(position = 'top', clear = FALSE),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE,
    pageLength = 15
  )
) %>%
  formatStyle(columns = c('clade_coverage', 'subclade_coverage'), 
              backgroundColor = styleInterval(c(50, 80), c('lightcoral', 'lightyellow', 'lightgreen')))
```

## Genome Completeness 
```{r}
# Table 3: Genome completeness summary by subtype (incorporating type information)
# Table 3: Genome completeness summary by subtype (incorporating type information)
genome_completeness_summary <- data %>%
  mutate(
    # Create a more informative subtype category
    subtype_category = case_when(
      subtype %in% c("H1N1", "H3N2", "Victoria") ~ subtype,
      type == "InfluenzaA" & !subtype %in% c("H1N1", "H3N2") ~ paste0("InfluenzaA_", subtype),
      type == "InfluenzaB" & subtype != "Victoria" ~ paste0("InfluenzaB_", subtype),
      is.na(subtype) | subtype == "" ~ paste0(type, "_unknown"),
      TRUE ~ paste0(type, "_", subtype)
    ),
    # Subtyping completeness
    complete_subtype = case_when(
      type == "InfluenzaA" & subtype %in% c("H1N1", "H3N2") ~ "Complete",
      type == "InfluenzaB" & subtype == "Victoria" ~ "Complete", 
      TRUE ~ "Incomplete"
    )
  ) %>%
  group_by(type, subtype_category, complete_subtype) %>%
  summarise(
    total_samples = n(),
    
    # Segment count statistics
    mean_segment_count = round(mean(segment_count, na.rm = TRUE), 1),
    median_segment_count = median(segment_count, na.rm = TRUE),
    min_segment_count = min(segment_count, na.rm = TRUE),
    max_segment_count = max(segment_count, na.rm = TRUE),
    
    # HA presence
    has_ha_count = sum(has_ha == "yes", na.rm = TRUE),
    has_ha_percent = round((has_ha_count / total_samples) * 100, 1),
    
    # HA and NA presence
    has_ha_and_na_count = sum(has_ha_and_na == "yes", na.rm = TRUE),
    has_ha_and_na_percent = round((has_ha_and_na_count / total_samples) * 100, 1),
    
    # Complete genome (8 segments)
    complete_genome_count = sum(segment_count == 8, na.rm = TRUE),
    complete_genome_percent = round((complete_genome_count / total_samples) * 100, 1),
    
    .groups = 'drop'
  ) %>%
  arrange(type, desc(total_samples))

datatable(
  genome_completeness_summary,
  caption = "Genome Completeness Summary by Type and Subtype",
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE
  )
) %>%
  formatStyle(columns = c('has_ha_percent', 'has_ha_and_na_percent', 'complete_genome_percent'), 
              backgroundColor = styleInterval(c(50, 80), c('lightcoral', 'lightyellow', 'lightgreen'))) %>%
  formatStyle(columns = 'complete_subtype',
              backgroundColor = styleEqual(c('Complete', 'Incomplete'), c('lightgreen', 'lightcoral'))) %>%
  formatRound(columns = c('mean_segment_count'), digits = 1)

```

## Interactive Table of all frequency data
```{r}
#| column: screen-inset-shaded
library(tidyverse)
library(DT)

datatable(
  data, 
  extensions = 'Buttons', 
  filter = list(position = 'top', clear = FALSE),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE # Enables horizontal scrolling
  )
) 
```

```{r}
# Table 3: Genome completeness summary by subtype (incorporating type information)
genome_completeness_summary <- data %>%
  mutate(
    # Create a more informative subtype category
    subtype_category = case_when(
      subtype %in% c("H1N1", "H3N2", "Victoria") ~ subtype,
      type == "InfluenzaA" & !subtype %in% c("H1N1", "H3N2") ~ paste0("InfluenzaA_", subtype),
      type == "InfluenzaB" & subtype != "Victoria" ~ paste0("InfluenzaB_", subtype),
      is.na(subtype) | subtype == "" ~ paste0(type, "_unknown"),
      TRUE ~ paste0(type, "_", subtype)
    ),
    # Subtyping completeness
    complete_subtype = case_when(
      type == "InfluenzaA" & subtype %in% c("H1N1", "H3N2") ~ "Complete",
      type == "InfluenzaB" & subtype == "Victoria" ~ "Complete", 
      TRUE ~ "Incomplete"
    )
  ) %>%
  group_by(type, subtype_category, complete_subtype) %>%
  summarise(
    total_samples = n(),
    
    # Segment count statistics
    mean_segment_count = round(mean(segment_count, na.rm = TRUE), 1),
    median_segment_count = median(segment_count, na.rm = TRUE),
    min_segment_count = min(segment_count, na.rm = TRUE),
    max_segment_count = max(segment_count, na.rm = TRUE),
    
    # HA presence
    has_ha_count = sum(has_ha == "yes", na.rm = TRUE),
    has_ha_percent = round((has_ha_count / total_samples) * 100, 1),
    
    # HA and NA presence
    has_ha_and_na_count = sum(has_ha_and_na == "yes", na.rm = TRUE),
    has_ha_and_na_percent = round((has_ha_and_na_count / total_samples) * 100, 1),
    
    # Complete genome (8 segments)
    complete_genome_count = sum(segment_count == 8, na.rm = TRUE),
    complete_genome_percent = round((complete_genome_count / total_samples) * 100, 1),
    
    .groups = 'drop'
  ) %>%
  arrange(type, desc(total_samples))

datatable(
  genome_completeness_summary,
  caption = "Genome Completeness Summary by Type and Subtype",
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'print'),
    scrollX = TRUE
  ),
  colnames = c(
    'Type' = 'type',
    'Subtype' = 'subtype_category',
    'Status' = 'complete_subtype',
    'N' = 'total_samples',
    'Mean Seg' = 'mean_segment_count',
    'Med Seg' = 'median_segment_count',
    'Min Seg' = 'min_segment_count',
    'Max Seg' = 'max_segment_count',
    'HA (n)' = 'has_ha_count',
    'HA (%)' = 'has_ha_percent',
    'HA+NA (n)' = 'has_ha_and_na_count',
    'HA+NA (%)' = 'has_ha_and_na_percent',
    'Complete (n)' = 'complete_genome_count',
    'Complete (%)' = 'complete_genome_percent'
  )
) %>%
  formatStyle(columns = c('HA (%)', 'HA+NA (%)', 'Complete (%)'), 
              backgroundColor = styleInterval(c(50, 80), c('lightcoral', 'lightyellow', 'lightgreen'))) %>%
  formatStyle(columns = 'Status',
              backgroundColor = styleEqual(c('Complete', 'Incomplete'), c('lightgreen', 'lightcoral'))) %>%
  formatRound(columns = c('Mean Seg'), digits = 1)
```

# Roadmap 

- Natural language information at the beggining of the report outlining
  - the current date range captured in this report
  - Is influenza a or b dominating 
  - which subtype is dominating 
- Table breakdown of influenza subtype and clade/subclade numbers 
-- Proceed with tables and plots -- 
- Dataset attributes
  - Interactive heatmap of samples with the number of present segments availible. 

## Features 
- Automatic color selection for each clade and subclade within subtypes.  
  